Competing Constructions for German Measure Phrases

### 2017-03-16 Latex/article.tex
work on previous analyses

Turning to \ref{it:intro:noifl}, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent'.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
A non-inflected variant \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected variant \textit{zwei Gläser roter Wein} co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
The statistical models presented in Section~\ref{sec:corpusmodels} will, however, show whether they (or other highly grammaticalised measure nouns) favor the NAC or the PGC through the estimation of lemma-specific random intercepts.

@@

This section reviews some existing analyses of the PGC-NAC alternation or related issues.
I do \textit{not} review any of the semantics literature on partitives vs.\ pseudo-partitives because the constructions of interest here are all semantically pseudo-partitives, and thus none of the variance in the choice of the PGC and NAC can be expected to be explained by a discussion along these lines.
The same is mostly true for the work on the grammaticalisation paths leading to the formation of pseudo-partitives (\eg\ \citealp{Brems2003,DeclerckBrems2016,Koptjevskaja2001,Rutkowski2007}).
It is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).%
\footnote{Notice that the description of German in \cite[547--549]{Koptjevskaja2001} (mostly referring \citealp{Eschenbach1993}) is partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
Her example (29b) from \cite[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun.

@@

It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \ref{page:femininesyncretism} and an ambiguity of the preposition \textit{troz}.
Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
}

% Syntactic structure: headedness (Galmann>Löbel, Eschenbach p. 213, Vos: DPC/DCC), difference to English (Brems)
% Syntactic structure: adjectives-as-determiners and layers in NP, role assignment to nouns in PGC/NAC
%     Gallmann>Bhatt1990 DP[gen] vs NP[agree]
% Influencing factors: Hentschel, Zimmer
% Squeeze in Anttila & Fong

@@

@@

@@

\subsection{The right statistics}
\label{ssec:rightstatistics}

>>>>

Turning to \ref{it:intro:noifl}, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent'.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
A non-inflected variant \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected variant \textit{zwei Gläser roter Wein} co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
See also Section~\ref{sec:analyses}.

@@

This section reviews some existing analyses of the PGC-NAC alternation or related issues.
I do \textit{not} review any of the semantics literature on partitives vs.\ pseudo-partitives because the constructions of interest here are all semantically pseudo-partitives, and thus none of the variance in the choice of the PGC and NAC can be expected to be explained by a discussion along these lines.
The same is mostly true for the work on the grammaticalisation of pseudo-partitives and measure noun constructions (\eg\ \citealp{Brems2003,DeclerckBrems2016,Koptjevskaja2001,Rutkowski2007}).
It is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).%
\footnote{Notice that the description of German in \cite[547--549]{Koptjevskaja2001} (mostly referring \citealp{Eschenbach1993}) is partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
Her example (29b) from \cite[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun.

@@

It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \ref{page:femininesyncretism} and an ambiguity of the preposition \textit{troz}.
Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
}
Since the alternating constructions are both clearly pseudo-partitives and not in any way partitives, this diachronic development (now completed) is not expected to have any effect on the alternation.
More pertinent are approaches like \cite{Brems2003,DeclerckBrems2016}.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns become quantifiers instead of nouns, leading to a reanalysis of the NP wherein the kind noun becomes the head.
This is supported by convincing syntactic evidence, for example subject-verb agreement.
In German MNPs, the measure noun remains the head, however (see immediately below), and I know of no strong hypotheses stating that the degree of grammaticalisation of the measure noun influences the alternation in any way.
The statistical models presented in Section~\ref{sec:corpusmodels} could in principle show, however, whether they (or other highly grammaticalised measure nouns) favor the NAC or the PGC through the estimation of lemma-specific random intercepts.

@@

% Syntactic structure: headedness (Galmann>Löbel, Eschenbach p. 213, Vos: DPC/DCC), difference to English (Brems)
% Syntactic structure: adjectives-as-determiners and layers in NP, role assignment to nouns in PGC/NAC
%     Gallmann>Bhatt1990 DP[gen] vs NP[agree]
%     Mention problem of no case assigner for NAC, so sub-syntactic process or theta theory wrong
% Influencing factors: Hentschel, Zimmer
% Squeeze in Anttila & Fong

@@

% Mention: Corpus-as-input is OK as a hypothesis but NOT as a research hypothesis
%  => otherwise any spurious pattern found in corpora (meeting certain criteria)
%  => is automatically "cognitive"

@@

\subsection{There are no right statistics!}
\label{ssec:rightstatistics}
